Running an open experiment: transparency and reproducibility in soil and ecosystem

### 2016-02-26 manuscript_openexperiment/Open_experiment.Rmd
Incorporated Peyton's comments

---
title: 'Running an open experiment: transparency and reproducibility in soil and ecosystem
  science'
author: "Ben Bond-Lamberty^1^, A. Peyton Smith^2^, Vanessa Bailey^2^"
date: "February 2, 2016"
output: word_document
---
^1^Joint Global Change Research Institute, DOE Pacific Northwest National Laboratory, College Park, MD USA; 
bondlamberty@pnnl.gov

@@

Researchers in soil and ecosystem science, and almost every other field, as increasingly being pushed--by funders, journals, governments, and themselves--to ensure transparency and reproducibility of their work. A key part of this effort is a move towards *open data* as a way to fight post-publication data loss, improve data and code quality, and enable powerful meta- and cross-disciplinary analyses. Here we describe a recent "open experiment", in which we documented every aspect of a soil incubation experiment online. This involved making raw data, scripts, diagnostics, final analyses, and manuscripts publicly available in real time. We found that using tools such as version control, issue tracking, and open-source statistical software improved data integrity, accelerated our team's communications, and ensured transparency. There are many avenues to improve scientific reproducibility and data availability, of which is this only one example, and it is not an approach suited for every experiment or situation. Nonetheless, we encourage the communities in our respective fields to consider its advantages, and to lead rather than follow with respect to scientific reproducibility, transparency, and data availability.

@@

Science is becomingly increasingly collaborative and data-intensive {Adams, 2012 #4348}. At the same time, a wide variety of factors are encouraging, at times pushing, scientists to increase data access and use "best practices" in dealing with data and code {Hart, 2015 #4292;Wilson, 2014 #4337}.

I really want to put something in about research collaborations becoming increasingly global, and the current state of the internet and associated collaborative tools create opportunities never-before-known to conduct research across (previously) geographic barriers.

Journals are adopting increasingly stringent data access and deposition policies, e.g. Scientific Data[^1], PLOS ONE[^2], and Science[^3]. These policies generally share common assumptions and goals: maximizing access to scientific data; encouraging deposition into structured repositories as opposed to journal supplementary information; and specifying that it is not acceptable for authors to be solely responsible for ensuring data access. Funders are moving in this direction as well, with organizations such as the U.S. National Science Foundation[^4] and the U.K. Wellcome Trust[^5], along with many others, requiring explicit data management plans, unfettered reasonable access to primary data, and use of established repositories.

@@

Finally, growing numbers of scientists are pushing for open science and data on moral/political grounds, as well as purely scientific ones, in particular ensuring reproducibility {Stodden, 2011 #4340} and enabling larger synthetic activities. Such analyses {Wolkovich, 2012 #3303} are made possible by the assembly of large, internally consistent data sets. Examples in ecology, soil science, and biogeosciences include BAAD {Falster, 2015 #4209}, TRY {Kattge, 2011 #3064}, FLUXNET {Baldocchi, 2001 #2883}, and SRDB {Bond-Lamberty, 2010 #2320}. Online databases and collaborative tools have expanded the abilities of researchers to collaborate across large distances, both improving data access and facilitating research partnerships (TODO).

@@

**Reproducibility**

Reproducibility of experimental results is at the heart of science {Kuhn, 1962 #4341}. Too often, however, sufficient details and data are not publicly available to even attempt to reproduce a study. For example, surveying the 2000-2014 biomedical literature, Iqbal et al. {, 2016 #4334} found that none of 441 randomly-chosen studies provided raw data, and only one provided full protocols. The issue of reproducibility in ecology (and its many related fields) has been raised and debated for years {Cassey, 2006 #4349} but ecological, soil science, and global change journals differ widely in their data deposition requirements.

As code and data have become increasingly intertwined, the availability of the former has become a fundamental problem as well. Modeling groups have worked to improve reproducibility and archiving practices {Thornton, 2005 #2310}. Scientists in all fields are increasingly building and using software in their work, however, often without strong training in this area {Wilson, 2014 #4337}. In addition to open-source data analysis languages such as R {R Development Core Team, 2016 #4314}, scientific workflow systems such as Kepler (https://kepler-project.org) or Taverna (http://www.taverna.org.uk) record information about the data processing, analytical process and decisions, and statistical analysis. Providing open code does not magically produce bug-free code, or mistake free analyses {Easterbrook, 2014 #4137}. But it does encourage authors to invest the time upfront to clean up their code, data, and documentation when a paper is written, rather than deferring this task (often, forever).

@@

Data loss hits ecosystem, soil, and global change ecology particularly hard, as climate changes make ecological data effectively irreproducible {Wolkovich, 2012 #3303}: we can never remeasure exactly the same system state. A subtler data loss issue is the "file drawer problem" {Rosenthal, 1979 #2291}, where unpublished but potentially valuable data are lost forever. Scientists' use of strong and consistent data curation practices {Hart, 2015 #4292}{Rüegg, 2014 #3886} can mitigate the problem, but all our anecdotal experience and quantitative studies {Vines, 2014 #4130} suggest that in the long term, data cannot be reliably preserved by individual researchers. This in turns means that the role of established, structured data and code repositories is critical.

TODO - VLB: Supplementary Information sections have become repositories of data and of non-core analyses. How does our open experiment replace (improve upon?) SI? Thoughts?  - journals not ready, and don't want {Cassey, 2006 #4349}.

@@

@@

* The entire data processing and analytical system is online and documented. It is written in R {R Development Core Team, 2016 #4314}, an open-source and widely used language and environment for statistical computing and graphics.
* The version control system let us make incremental changes, work out problems, look at histories (i.e., who made what change when).
* An issue tracker let us discuss problems, reference changes in the repository, create to-do lists, and assign responsibilities.
* Information webpage on the experiment and broader project.
* Manuscripts as well! (Including this one.) Results are linked directly into the manuscripts, so that if a result changes, it is updated automatically in the text. Tools such as R Markdown[^7] have made this process much easier to build and maintain.

@@

@@

Transparency is no magic panacea ensuring reproducibility {Easterbrook, 2014 #4137}, and may raise genuine concerns about, for example, protecting scientists from harassment {Lewandowsky, 2016 #4342}. Scientists also frequently cite other concerns about data sharing, for example being "scooped" not receiving sufficient credit, and time constraints {Wolkovich, 2012 #3303}. We cannot easily dismiss of these: for example, does our community adequately credit researchers who contribute to global databases that subsequently produce high-impact papers? Such meta-analyses rely on the collection of primary data, and it is critical that field and experimental researchers efforts are adequately valued and cited {Kueffer, 2011 #4335}. We also recognize that, in general, our professional and career incentives do not yet align well with "an open research culture" {Nosek, 2015 #4343}--that is, scientists do not receive 'proper' credit for datasets, relative to publications, even when the former may be much more broadly valuable.

The model we present here--a public repository updated throughout the experiment, analysis, and publication process--is far from perfect. First, there are a number of "best practices" of scientific computing {Wilson, 2014 #4337} that we did not employ: for example, there was no unit testing to ensure module correctness, nor an automated build tool automating the workflow (i.e., the individual scripts all have to be run manually). Second, this approach may not be appropriate for, nor applicable to, all study types, even just in our professional fields of soil, ecosystem, and global change science; it certainly is not appropriate for sensitive human-subject data. Third, Git and GitHub are designed for working with code, not data, although this situation continues to improve on a technical level[^6].

@@

@@

>>>>

---
title: 'Running an open experiment: transparency and reproducibility in soil and ecosystem
  science'
author: "Ben Bond-Lamberty^1^, A. Peyton Smith^2^, Vanessa Bailey^2^"
date: "February 2, 2016"
output: html_document
---
^1^Joint Global Change Research Institute, DOE Pacific Northwest National Laboratory, College Park, MD USA; 
bondlamberty@pnnl.gov

@@

Researchers in soil and ecosystem science, and almost every other field, as increasingly being pushed--by funders, journals, governments, and themselves--to increase transparency and ensure reproducibility of their work. A key part of this effort is a move towards *open data* as a way to fight post-publication data loss, improve data and code quality, and enable powerful meta- and cross-disciplinary analyses. Here we describe a recent "open experiment", in which we documented every aspect of a soil incubation experiment online. This involved making all raw data, scripts, diagnostics, final analyses, and manuscripts publicly available in real time. We found that using tools such as version control, issue tracking, and open-source statistical software improved data integrity, accelerated our team's communication and productivity, and ensured transparency. There are many avenues to improve scientific reproducibility and data availability, of which is this only one example, and it is not an approach suited for every experiment or situation. Nonetheless, we encourage the communities in our respective fields to consider its advantages, and to lead rather than follow with respect to scientific reproducibility, transparency, and data availability.

@@

Science is becomingly increasingly collaborative and data-intensive {Adams, 2012 #4348}; in conjunction with revolutions in Internet-based communication, this has created new reserach opportunities across former geographic and disciplinary barriers. At the same time, a wide variety of factors are encouraging, at times pushing, scientists to increase data access and use "best practices" in dealing with data and code {Hart, 2015 #4292;Wilson, 2014 #4337}.

These factors include journals, which are adopting increasingly stringent data access and deposition policies, e.g. Scientific Data[^1], PLOS ONE[^2], and Science[^3]. These policies generally share common assumptions and goals: maximizing access to scientific data; encouraging deposition into structured repositories as opposed to journal supplementary information; and specifying that it is not acceptable for authors to be solely responsible for ensuring data access. Funders are moving in this direction as well, with organizations such as the U.S. National Science Foundation[^4] and the U.K. Wellcome Trust[^5], along with many others, requiring explicit data management plans, unfettered reasonable access to primary data, and use of established repositories.

@@

Finally, growing numbers of scientists are pushing for open science and data on moral/political grounds, with respect to taxpayer-funded research, as well as purely scientific ones, in particular ensuring reproducibility {Stodden, 2011 #4340} and enabling larger synthetic activities. Such analyses {Wolkovich, 2012 #3303} are made possible by the assembly of large, internally consistent data sets. Examples in ecology, soil science, and biogeosciences include BAAD {Falster, 2015 #4209}, TRY {Kattge, 2011 #3064}, FLUXNET {Baldocchi, 2001 #2883}, and SRDB {Bond-Lamberty, 2010 #2320}. Online databases and collaborative tools have expanded the abilities of researchers to collaborate across large distances, both improving data access and facilitating research partnerships (TODO).

@@

**Repeatability and reproducibility**

Reproducibility of experimental results is at the heart of science and a requirement for results to be accepted as factual {Kuhn, 1962 #4341}. Too often, however, sufficient details and data are not publicly available to even attempt to repeat a study (i.e. perform it again in a comparable manner, while not expecting exactly the same result). For example, surveying the 2000-2014 biomedical literature, Iqbal et al. {, 2016 #4334} found that none of 441 randomly-chosen studies provided raw data, and only one provided full protocols. The issues of reproducibility and repeatability in ecology (and its many related fields) have been raised and debated for years {Cassey, 2006 #4349}{Ellison, 2010 #4363} but ecological, soil science, and global change journals differ widely in their data deposition requirements, when such requirements exist at all.

As code and data have become increasingly intertwined, the availability of the former has become a fundamental problem as well. Modeling groups have worked to improve reproducibility and archiving practices {Thornton, 2005 #2310}. Scientists in all fields are increasingly building and using software in their work, though often without strong training in this area {Wilson, 2014 #4337}. In addition to open-source data analysis languages such as R {R Development Core Team, 2016 #4314}, scientific workflow systems such as Kepler (https://kepler-project.org) or Taverna (http://www.taverna.org.uk) record information about the data processing, analytical process and decisions, and statistical analysis. Providing open code does not magically produce bug-free code, mistake-free analyses, or instantly better science {Easterbrook, 2014 #4137}. But it does encourage authors to invest the time upfront to clean up their code, data, and documentation when a paper is written, rather than deferring this task (often until key details have been forgotten, if not forever).

@@

Data loss hits ecosystem, soil, and global change ecology particularly hard, as climate changes make ecological data effectively irreproducible {Wolkovich, 2012 #3303}: we can never remeasure exactly the same system state. This results in a critical role for syntheses and meta-analyses {Ellison, 2010 #4363}{Cassey, 2006 #4349}, which depend on the existence of documented protocols and (ideally raw) data. A subtler data loss issue is the "file drawer problem" {Rosenthal, 1979 #2291}, where unpublished but potentially valuable data are lost forever. Scientists' use of strong and consistent data curation practices {Hart, 2015 #4292}{Rüegg, 2014 #3886} can mitigate the problem, but all our anecdotal experience and quantitative studies {Vines, 2014 #4130} suggest that in the long term, data cannot be reliably preserved by individual researchers.

This in turns means that the role of established, structured data and code repositories is critical.

TODO - VLB: Supplementary Information sections have become repositories of data and of non-core analyses. Yet, data provided in Supplemental Information sections while available for use, may be accessible only in inconvenient formats (e.g. pdf) or have restrictions for use in many journals (Molloy 2011, Neylon 2012).

How does our open experiment replace (improve upon?) SI? Thoughts?  - journals not ready, and don't want {Cassey, 2006 #4349}.

@@

While open science is often touted as paving the way for enhancing public trust in science, there are still many social and political challenges to overcome (Grandy et al. 2012). For example, the language used in academic publications is often difficult for someone outside the specific field to understand (whether the reader is a scientist or not) and so simply providing unfettered access of scientific publications to the public may not improve the public’s perception and trust in science. While there is a strong need for the public to better trust science, the reverse is also true.  Trust by scientists and institutions in the public to accurately interpret results scientists and institutions is also another challenge in the quest for open science (CITE). List examples: BSE risks, causes and beef sales (90s), or  GMO perceptions vs science (described in Millstone & van Zwanenberg 2000 Nature Medicine), or public perception fueling debates on bird flu mutation research which resulted in 60 day ban on bird flu research…hmm, perhaps not the best examples.

@@

* The entire data processing and analytical system is online and documented. It is written in R {R Development Core Team, 2016 #4314}, an open-source and widely used language and environment for statistical computing and graphics.
* The version control system let us make incremental changes, work out problems, look at histories (i.e., who made what change when).
* An issue tracker let us discuss problems, reference changes in the repository, create to-do lists, and assign responsibilities.
* Informational webpage on the experiment and broader project.
* Manuscripts as well! (Including this one.) Results are linked directly into the manuscripts, so that if a result changes, it is updated automatically in the text. Tools such as R Markdown[^7] have made this process much easier to build and maintain.
* Log files provide an audit trail {Ellison, 2010 #4363} of what analytical steps were taken and, critically, the specific versions used for each unit of software, including the main R system itself. This is critical as software changes with time.

@@

Disadvantages:
•	Convincing co-authors who are uncomfortable with pre-published data being made publically available of the benefits of open science
•	Having Peyton adhere to a level of professionalism in all comments and communications as they were made publically available ☺ #firstworldproblem
•	Writing a manuscript in code was not a skill all co-authors possessed, nor was it the most easy or efficient way to collaborate on a paper (ex: reference manager?) ☺

@@

Transparency is no magic panacea ensuring reproducibility {Easterbrook, 2014 #4137}, and may raise genuine concerns about, for example, protecting scientists from harassment {Lewandowsky, 2016 #4342}. Scientists also frequently cite other concerns about data sharing, for example being "scooped" not receiving sufficient credit, and time constraints {Wolkovich, 2012 #3303}. We cannot easily dismiss of these: for example, does our community adequately credit researchers who contribute to global databases that subsequently produce high-impact papers? Such meta-analyses rely on the collection of primary data, and it is critical that field and experimental researchers efforts are adequately valued and cited {Kueffer, 2011 #4335}. We also recognize that, in general, our professional and career incentives do not yet align well with "an open research culture" {Nosek, 2015 #4343}. That is, scientists do not receive 'proper' credit for datasets relative to publications, even when the former may be much more broadly valuable, in almost every area: hiring, promotion and tenure decisions, impact metrics, etc.

The model we present here--a public repository updated throughout the experiment, analysis, and publication process--is far from perfect. First, there are a number of "best practices" of scientific computing {Wilson, 2014 #4337} that we did not employ: for example, there was no unit testing to ensure module correctness, nor a build tool automating the workflow (i.e., the individual scripts all have to be run manually). Second, this approach may not be appropriate for, nor applicable to, all study types, even just in our professional fields of soil, ecosystem, and global change science; it certainly is not appropriate for sensitive human-subject data. Third, Git and GitHub are designed for working with code, not data, although this situation continues to improve on a technical level[^6].

@@

There have been attempts in environmental science to create data repositories – for example, the Data Observation Network for Earth, DataONE (https://www.dataone.org/what-dataone), provides a repository for earth science data and makes it accessible to the both the scientific and general public with the mission to: “ensure the preservation, access, use and reuse of multi-scale, multi-discipline, and multi-national science data…[for use] by researchers, educators, and the public to better understand and conserve life on earth and the environment that sustains it”  and allows researchers to cite data, workflows or visualizations located in the DataONE archive (Reichman et al. 2011). Nonetheless (TODO: it's a lot better than nothing, and offers instrument-to-final product reproducibility).

Another nonetheless: A study from the UK group, Research Information Network (as mentioned in Molloy 2011), titled “To share or not to share: research data outputs”[^7], interviewed 100 scientists about open data and in their summary of findings, study participants reported that they would be more incentivized to partake in open data practices if there was more “evidence that there are benefits to be had from publishing datasets (e.g. through case studies)”[^8]. So in essence, making our open science experience into a case study may help contribute to the growing movement of open science/may encourage other researchers to participate.

[^7]: http://www.rin.ac.uk/our-work/data-management-and-curation/share-or-not-share-research-data-outputs
[^8]: http://www.rin.ac.uk/system/files/attachments/To-share-data-outputs-report.pdf

@@

APS: Something that we haven’t really addressed in this paper is what I consider as one of the most valuable advantages of having our experiment open: The exposure! More visibility for us as scientists and PNNL.  I love that. I also think that (while it hasn’t yet), one of us will get on another publication/collaboration due to this dataset being made public.  I’m not sure where that goes in this MS and if there is a citation to back up my claims – but I’ll look for one nevertheless.
